Recipient plant traits are important determinants of the impacts of heterospecific pollen upon plant reproduction

### 2019-03-15 Manuscript_draft/Manuscript_draft.tex
edits legend

\documentclass[11pt,a4paper]{article}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs}
\usepackage{graphicx}
% Redefine \includegraphics so that, unless explicit options are
% given, the image width will not exceed the width of the page.
% Images get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\ScaleIfNeeded{%
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother
\let\Oldincludegraphics\includegraphics
{%
 \catcode`\@=11\relax%
 \gdef\includegraphics{\@ifnextchar[{\Oldincludegraphics}{\Oldincludegraphics[width=\ScaleIfNeeded]}}%
}%
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={working title Compatibility system and stigma size are the main predictors of heterospecific pollen effect},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

@@

\title{working title Compatibility system and stigma size are the main
predictors of heterospecific pollen effect}
\author{
Jose B. Lanuza
Ignasi Bartomeus
Tia-Lynn Ashman
Romina Rader
}
\date{}
% Allowing for landscape pages
\usepackage{lscape}
\newcommand{\blandscape}{\begin{landscape}}
\newcommand{\elandscape}{\end{landscape}}

@@

\begin{document}
%%Edited by @jhollist 09/15/2014
%%Adds title from YAML
\begin{singlespace}
\begin{center}
\huge working title Compatibility system and stigma size are the main
predictors of heterospecific pollen effect
\end{center}
%%Adds Author, correspond email asterisk, and affilnum from YAML
\begin{center}
\large
Jose B. Lanuza \textsuperscript{*} \textsuperscript{1} 
Ignasi Bartomeus \textsuperscript{2} 
Tia-Lynn Ashman \textsuperscript{3} 
Romina Rader \textsuperscript{1} 
\end{center}
%%Adds affiliations from YAML
\begin{justify}
\footnotesize \emph{ 
\\*
\textsuperscript{1}University of New England (Australia)\\*
\\*
\textsuperscript{2}Estacion Biologica de Donana (EBD-CSIC), E-41092 Sevilla, Spain\\*
\\*
\textsuperscript{3}Department of Biological Sciences, University of Pittsburgh 4249 Fifth
Avenue, Pittsburgh, Pennsylvania 15260-3929 USA\\*
}
%%Adds corresponding author email(s) from YAML
\newcounter{num}
\setcounter{num}{1}
\\[0.1cm]
\footnotesize \emph{ 
\ifnum\value{num}=1%
\textsuperscript{*} corresponding author:
\fi
\href{mailto:barragansljose@gmail.com}{\nolinkurl{barragansljose@gmail.com}}
\stepcounter{num}
}
\end{justify}
%%Adds date from YAML
\normalsize

@@

Pollinator sharing can have negative consequences for plant fitness with
the arrival of foreign pollen. However, the costs of heterospecific
pollen are not yet well understood. We conducted a glasshouse experiment
to understand how phylogenetic relatedness and plant traits mediate the
impacts of heterospecific pollen transfer. We conducted 1800 crosses by
experimentally transferring pollen (50\% and 100\% ratio) with
reciprocal crosses between 10 species belonging to three different
families: Brassicaceae, Solanaceae and Convolvulaceae. Seed set was used
as proxy of plant fitness. We found that for 65\% of the treatments with
50\% mix reduced seed set. Moreover, the reduction in seed set was
dependent on the degree of relatedness and reproductive traits of the
pollen recipient and not the pollen donor. Our results show that certain
traits, particularly compatibility system, are critical in understanding
the costs of heterospecific pollen. \vspace{3mm}\hrule

@@

In most ecosystems, plant species normally coexist and share their
floral visitors with other species Waser et al. (1996). From the plants'
perspective, pollinator sharing can be positive for some plants
Carvalheiro et al. (2014) or negative for others Pauw (2013), depending
on the facilitation gradient. An increasing number of visits often
correlates with higher chances of fertilization Engel and Irwin (2003).
However this is not always the case, among these possible flower
visitors there are also nectar robbers and pollen thieves Inouye (1980);
Magrach et al. (2017). Receiving both sufficient quantity and quality
deposited on the stigma is thus highly relevant to the pollination
success of the plant Aizen and Harder (2007).

By visiting many plant species, many pollinators are responsible for
conspecific pollen loss and the transport of foreign pollen, both of
which can have important detrimental effects on species fitness Morales
and Traveset (2008); Ashman and Arceo-Gómez (2013); Arceo-Gómez and
Ashman (2016). Foreign pollen arrival can play an important role in
plant species fitness but outcomes are variable and appear to be context
dependent as there is not always a decrease in fitness Morales and
Traveset (2008). Some of this variation is likely due to the enormous
variability of foreign pollen transferred across systems ranging from 0
to 75 percent. However, most studies report ranges of heterospecific
pollen between 0 and 20 percent of the total pollen load Bartomeus et
al. (2008) Montgomery and Rathcke (2012); Ashman and Arceo-Gómez (2013);
Fang and Huang (2013), yet even these relatively low amounts of
heterospecific pollen transferred can decrease fitness greatly Thomson
et al. (1982). While we now have some understanding of the impacts of
heterospecific pollen quantity, we have less understanding of other
factors that could be driving the variation in impacts upon fitness.
Ashman and Arceo-Gómez (2013) postulated the first predictive framework
that identifies a need to understand how plant traits might mediate
heterospecific pollen effect, whereby mating system and pollen size were
predicted to potentially mediate the impact of foreign pollen transfer
on plant fitness. This concept is supported by specific case studies,
such as Tong and Huang (2016) that demonstrate an asymmetrical effect in
6 species of \emph{Pedicularis} whereby the pollen of long styled
species was able to grow the full length of the style on short styled
species but not vice versa. While this suggests that the impacts of
heterospecific pollen may differ among pollen donor and recipient, few
studies have been conducted to ascertain whether this pattern is in fact
a general trend or to identify the extent to which other plant traits
are critical to heterospecific pollen impacts.

Plant traits are crucial to understand heterospecific pollen effect but
the multifactorial nature of the traits that are involve in the
pollen-pistil interaction make difficult to unravel what are the main
traits in driving the effect. These traits can be seen from a male
perspective of both donor and recipient where pollen size, pollen
aperture number and pollen allelopathy are key components to understand
the outcome of foreign pollen arrival Murphy and Aarssen (1995); Ashman
and Arceo-Gómez (2013). In Ashman and Arceo-Gómez (2013) small pollen is
predicted to cause a greater fitness decrease, although this can be true
there are also other possibilities to consider which can obscure a
predictive framework like big pollen can clogg small stigmas with fewer
pollen grains, bigger stigmas are less likely to be clogged by small
pollen grains and bigger pollen can outcompete smaller pollen grains due
to faster pollen tube growth rate Williams and Rouse (1990). Moreover,
to understand the different mechanical or chemical effects of pollen
also the female traits of the pollen recipient have to be considered,
from the literature these main traits are: stigma size, style length,
number of ovules, incompatibility system and also a structural trait
such as flower morphology Montgomery and Rathcke (2012); Ashman and
Arceo-Gómez (2013); Tong and Huang (2016). For example, greater
stigmatic area is positively correlated with greater heterospecific
pollen deposition Montgomery and Rathcke (2012) and therefore possibly
with an increase in negative effect. For species that are
self-incompatible the barriers towards heterospecific pollen are
stronger than self-compatible species Ashman and Arceo-Gómez (2013).
Nonetheless, an effect of foreign pollen is a bit obscured by the
variability within species, however species that are strong selfers or
strong outcrossers have less variablity in mating systems and
predictions of effect could be more realistic (see figure 1 from
Whitehead et al. (2018)). Although past research has progress in the
understanding of what traits can mediate the effect as we have shown
here, there are multiple traits involved and multiple possible scenarios
still to be explored empirically for a full understanding of the
importance of heterospecific pollen effect in nature.

For the understanding at what level or intensity the interference of
pollen can occur is important to consider the relatedness of the
interacting species. Closely related species are more likely to have
similar traits Letten and Cornwell (2015). The similarity in traits
between closely related species can lead to higher chances of ovule
usurpation/abortion Arceo-Gómez and Ashman (2011) and therefore studies
predict and show a greater negative effect of closely related species
Ashman and Arceo-Gómez (2013); Arceo-Gómez and Ashman (2016). Few
studies however, have focused on the impacts of heterospecific pollen on
fitness of distantly related species Galen and Gregory (1989); Neiland
and Wilcock (1999). Despite the fact that far related species are also
able to decrease species fitness (REFS). Yet, most insects and most
stigmas have been found to carry multiple species of foreign pollen with
little attention to degree of relatedness (Arceo-Gómez and Ashman
(2016); Fang and Huang (2013). Understanding the role of foreign pollen
from distantly related species thus deserves greater attention. The
relatedness of foreign pollen gives a first snapshoot of where the
pollen competition can occur and therefore could be a proxy of effect.
With that said, Arceo-Gómez and Ashman (2016) is the only work until our
knowledge which has proven a greater effect of close related species
through a meta-analysis but with low sample sizes and lack of
significance. Therefore, there is a need to study the effect of
heterospecific pollen of far and close related species at community
level beyond single pairwise interactions.

Heterospecific pollen studies in nature have the complexity added of
great environmental variability which can lead to confounding
interpretations. Moreover, the great diversity of floral structures and
small sizes of some of them such as Asteraceae species make tedious or
almost impossible the studies of heterospecific pollen on the field. For
this reason, we investigated how floral reproductive traits and
relatedness mediate the impact of heterospecific pollen by creating an
artificial co-flowering community in a glasshouse with 10 species
belonging to three different families with heterogeneous reproductive
traits. Our study tries to answer the following questions: (i) Do
heterospecific pollen reduce seed set? To what extent do (ii) floral
reproductive traits and (iii) relatedness, mediate the impacts of
heterospecific pollen on seed set.

@@

\hypertarget{ref-whitehead2018}{}
Whitehead, M. R., R. Lanfear, R. J. Mitchell, and J. D. Karron. 2018.
Plant mating systems often vary widely among populations. Frontiers in
Ecology and Evolution 6:38.

>>>>

\documentclass[11pt,a4paper]{article}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs}
\usepackage{graphicx}
% Redefine \includegraphics so that, unless explicit options are
% given, the image width will not exceed the width of the page.
% Images get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\ScaleIfNeeded{%
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother
\let\Oldincludegraphics\includegraphics
{%
 \catcode`\@=11\relax%
 \gdef\includegraphics{\@ifnextchar[{\Oldincludegraphics}{\Oldincludegraphics[width=\ScaleIfNeeded]}}%
}%
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Recipient plant traits are important determinants of the impacts of heterospecific pollen upon plant reproduction},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

@@

\title{Recipient plant traits are important determinants of the impacts of
heterospecific pollen upon plant reproduction}
\author{
Jose B. Lanuza
Ignasi Bartomeus
Tia-Lynn Ashman
Romina Rader
}
\date{}
% Allowing for landscape pages
\usepackage{lscape}
\newcommand{\blandscape}{\begin{landscape}}
\newcommand{\elandscape}{\end{landscape}}

@@

\begin{document}
%%Edited by @jhollist 09/15/2014
%%Adds title from YAML
\begin{singlespace}
\begin{center}
\huge Recipient plant traits are important determinants of the impacts of
heterospecific pollen upon plant reproduction
\end{center}
%%Adds Author, correspond email asterisk, and affilnum from YAML
\begin{center}
\large
Jose B. Lanuza \textsuperscript{*} \textsuperscript{1} 
Ignasi Bartomeus \textsuperscript{2} 
Tia-Lynn Ashman \textsuperscript{3} 
Romina Rader \textsuperscript{1} 
\end{center}
%%Adds affiliations from YAML
\begin{justify}
\footnotesize \emph{ 
\\*
\textsuperscript{1}University of New England (Australia)\\*
\\*
\textsuperscript{2}Estacion Biologica de Donana (EBD-CSIC), E-41092 Sevilla, Spain\\*
\\*
\textsuperscript{3}Department of Biological Sciences, University of Pittsburgh 4249 Fifth
Avenue, Pittsburgh, Pennsylvania 15260-3929 USA\\*
}
%%Adds corresponding author email(s) from YAML
\newcounter{num}
\setcounter{num}{1}
\\[0.1cm]
\footnotesize \emph{ 
\ifnum\value{num}=1%
\textsuperscript{*} corresponding author:
\fi
\href{mailto:barragansljose@gmail.com}{\nolinkurl{barragansljose@gmail.com}}
\stepcounter{num}
}
\end{justify}
%%Adds date from YAML
\normalsize

@@

Pollinator sharing can have negative consequences for plant fitness with
the arrival of foreign pollen. However, the mechanisms underlying the
variation in outcomes as a result of pollen contribution by different
plant species are not yet well understood. We conducted a glasshouse
experiment to understand how plant traits and phylogenetic relatedness
mediate the impacts of 15 heterospecific pollen transfer. We conducted
1800 reciprocal crosses by experimentally transferring pollen (50\% and
100\% foreign pollen ratio) between 10 species belonging to three
different families: Brassicaceae, Solanaceae and Convolvulaceae. Seed
set was used as proxy of plant fitness. In the treatments of 100\%
foreign pollen, we found reduced seed set in X\% of the treatments. In
the treatments of 50\% foreign pollen, we found reduced seed set in 65\%
of the treatments. Moreover, the reduction in seed set was dependent on
the reproductive traits of the pollen recipient, but not the pollen
donor or relatedness. Our results show that certain traits of recipient
plants, particularly compatibility system, are critical in understanding
the costs of heterospecific pollen.

\vspace{3mm}\hrule

@@

In most ecosystems, plant species normally coexist and share their
floral visitors with other species Waser et al. (1996); Carvalheiro et
al. (2014). From the plants' perspective, pollinator sharing can be
positive for some plants as an increasing number of visits often
correlates with higher chances of fertilization Engel and Irwin (2003).
Yet, among these possible flower visitors there are also nectar robbers,
pollen thieves Inouye (1980); Magrach et al. (2017), and inconstant
pollinators that transfer foreign pollen from other plants Pauw (2013).
By visiting many plant species, many pollinators are responsible for
conspecific pollen loss and the transport of foreign pollen, both of
which can have important detrimental effects on species fitness Morales
and Traveset (2008); Ashman and Arceo-Gómez (2013); Arceo-Gómez and
Ashman (2016). Receiving both sufficient quantity and quality deposited
on the stigma is thus highly relevant to the pollination success of the
plant Aizen and Harder (2007). Foreign pollen arrival can play an
important role in plant species fitness but outcomes are variable and
appear to be context dependent as there is not always a decrease in
fitness Morales and Traveset (2008). Some of this variation is likely
due to the enormous variability of foreign pollen transferred across
systems ranging from 0 to 75 percent. However, most studies report
ranges of heterospecific pollen between 0 and 20 percent of the total
pollen load Bartomeus et al. (2008) Montgomery and Rathcke (2012);
Ashman and Arceo-Gómez (2013); Fang and Huang (2013), yet even these
relatively low amounts of heterospecific pollen transferred can decrease
fitness greatly Thomson et al. (1982).

While we have some understanding of the impacts of heterospecific pollen
quantity, we have little knoweldge of the factors that could be driving
the variation in pollen quality upon fitness. Plant traits are crucial
to understand heterospecific pollen effect but the multifactorial nature
of the traits that are involved in the pollen-pistil interaction make
difficult to unravel exactly which traits are driving the effect. Ashman
and Arceo-Gómez (2013) postulated the first predictive framework that
identifies a need to understand how plant traits might mediate
heterospecific pollen effect, whereby mating system and pollen size were
predicted to potentially mediate the impact of foreign pollen transfer
on plant fitness.

The concept of trait driven mechanisms is not new and is supported by
system specific studies. Pollen size, pollen aperture number and pollen
allelopathy are thought to be key components in understanding the
outcome of foreign pollen arrival Murphy and Aarssen (1995); Ashman and
Arceo-Gómez (2013). For example, small pollen is predicted to decrease
plant fitness because XXXXXX. Yet, large pollen can outcompete smaller
pollen grains due to faster pollen tube growth rate Williams and Rouse
(1990). Hence, understanding the different mechanical or chemical
effects of pollen requires knowledge of the female traits of the pollen
recipient to also be considered Montgomery and Rathcke (2012); Ashman
and Arceo-Gómez (2013); Tong and Huang (2016). For example, greater
stigmatic area is positively correlated with greater heterospecific
pollen deposition Montgomery and Rathcke (2012) and therefore likely to
result in an greater negative effect upon plant fitness. Further,
species that are self-incompatible are thought to be more resistant to
the negative impacts of heterospecific pollen than self-compatible
species Ashman and Arceo-Gómez (2013).

When both donor and recipient traits are considered together, other
combinations of traits are also likely to impact plant fitness. For
example, large pollen grains could potentially clog small stigmas with
fewer pollen grains, and larger stigmas are less likely to be clogged by
small pollen grains. Yet, few studies have considered how effects might
differ among donor and recipient species. Tong and Huang (2016)
demonstrate an asymmetrical effect in 6 species of Pedicularis whereby
foreign pollen of long styled species was able to grow the full length
of the style on short styled species but not vice versa. While this
suggests that the impacts of heterospecific pollen may differ among
pollen donor and recipient, few studies have been conducted to ascertain
whether this pattern is in fact a general trend or to identify the
extent to which other plant traits are critical to heterospecific pollen
impacts.

It is challenging to identify general patterns with respect to the
mechanisms driving foreign pollen impacts as results are often obscured
by the variability within and among species. Closely related species are
more likely to have similar traits Letten and Cornwell (2015). The
similarity in traits between closely related species can lead tohigher
chances of ovule usurpation/abortion Arceo-Gómez and Ashman (2011)
hence, greater negative effects of HP pollen are thought to be a
associated with more closely related species Ashman and Arceo-Gómez
(2013); Arceo-Gómez and Ashman (2016). Few studies however, have focused
on the impacts of heterospecific pollen on fitness of distantly related
species Galen and Gregory (1989); Neiland and Wilcock (1999) and those
that have, often report low sample sizes and a lack of significance.
Therefore, there is a need to study the effect of heterospecific pollen
of far and close related species at community level beyond single
pairwise interactions. Given that pollen carried on many insects and
stigmas has been found to carry multiple species of foreign pollen with
little attention to degree of relatedness Arceo-Gómez and Ashman (2016);
Fang and Huang (2013). understanding the role of foreign pollen from
distantly related species thus deserves greater attention

We investigated how floral reproductive traits and relatedness mediate
the impact of heterospecific pollen by creating an artificial
co-flowering community in a glasshouse with 10 species belonging to
three different families with heterogeneous reproductive traits. Our
study addressed the following questions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  To what extent does the amount of foreign pollen applied to stigmas
  impact plant reproductive fitness (i.e.~50\% and 100\% foreign pollen
  ratio.
\item
  How do floral reproductive traits and plant relatedness mediate the
  impacts of heterospecific pollen on seed set.
\end{enumerate}

@@
